The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-11 paper/iphas-dr2.tex

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering the Galactic latitude range $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey area.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
and to a mean $5\sigma$-depth of 
21.2 ($r$), 20.0 ($i$) and 20.3 ($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the new 97-column source catalogue.
We show that the global calibration is accurate to
0.03~mag (rms) by validating against the Sloan Digital Sky Survey (SDSS),
and provide a series of recommended quality criteria
which can be used to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select candidate \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure of the disk of the Milky Way.
This large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights 
at the INT in La Palma --
aims to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century \citep[see][]{Kohoutek1999}. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
IPHAS can expand
the limited bright samples of Galactic emission line objects 
previously available into larger, deeper, and more statistically-robust samples that will 
better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the IPHAS Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
%new asymptotic giant branch stars \citep{Wright2008};
new classical Be stars \citep{Raddi2013};
along with discoveries of new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Corradi2011, Viironen2011}
and new supernova remnants \citep{Sabin2013}.

Over the years it has become apparent that the legacy of IPHAS 
will also reach beyond these traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$, $i$ broadband photometry 
alongside H$\alpha$ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey’s unique $(r-\ha,\ r-i)$ colour-colour
diagram has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband H$\alpha$ 
alongside $r, i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys.
For such science applications to succeed, however, 
it is vital that the imaging data are transformed 
into a homogeneously calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

When the initial data release was created in late 2007,
only $\sim$half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release 
that takes the coverage up to over 90 percent of the survey area 
and includes a uniform calibration.
In this work we aim to
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new global photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

In \S\ref{sec:observations} we start by recapitulating the key points
of the survey observing strategy.
In \S\ref{sec:reduction} we describe the data reduction
and quality control procedures.
In \S\ref{sec:calibration} we explain the uniform re-calibration
and test our results against the Sloan Digital Sky Survey (SDSS).
In \S\ref{sec:catalogue} we explain how the source catalogue was compiled.
In \S\ref{sec:discussion} we discuss the properties of the catalogue
and in \S\ref{sec:demonstration} we demonstrate
the scientific exploitation of the colour/magnitude diagrams.
Finally, in \S\ref{sec:dataaccess} we discuss access
to the data and source code,
and in \S\ref{sec:conclusions} we conclude and outline
our future ambition

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release, we briefly recall some key points.
IPHAS is a 1800~deg$^2$. imaging survey of the northern Galactic Plane, 
providing images and photometry in Sloan $r, i$ 
along with narrowband H$\alpha$. 
It is carried out using the WFC 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive CCD photometry
of point sources in the Galactic Plane at visible wavelengths, 
and does so down to a limiting magnitude of $\sim$20.
The IPHAS footprint spans a box 
of roughly 180 by 10 degrees, 
taking in the entire northern Galactic Plane 
at latitudes $-5^{\circ} < b < +5^{\circ}$ 
and longitudes $30^{\circ} < l < 215^{\circ}$.

The WFC is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29~deg$^2$.
To cover the entire Northern Plane with some overlap,
the survey area was divided into 7635 telescope pointings.
Each of these pointings is accompanied by an offset position
displaced by $+$5 arcmin in declination 
and $+$5 arcmin in right ascension,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks. 
The basic unit of observation hence
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
The survey hence contains 15\,270 fields grouped into 7\,635 field pairs.
To achieve the desired survey depth of 20th magnitude or fainter, 
the filter exposure times were set at 120 sec (narrowband \ha), 
30 sec ($r$) and 10 sec ($i$)
in the majority of the survey observations.\footnote{In 2003 
the $r$-band exposure time was 10~sec instead of 30~sec,
and since Oct 2010 the $i$-band exposure time 
has been increased from 10~sec to 20~sec.}

Data-taking began in the second half of 2003, 
and every field had been observed at least once by the end of 2008.
At that time only 76\% of the field pairs 
satisfied our minimum quality criteria however,  
often due to the effects of clouds, poor seeing, or technical faults
(the quality criteria will be detailed in the next section). 
Since then, a programme of repeat observations has been in place 
to improve data quality. 
As a result, 92\% of the survey 
now benefits from quality-approved data.
The most recent observations which are part of this release
were obtained in November 2012.

\begin{figure*}
        \includegraphics[width=1\linewidth]{./plots/footprint/footprint_small.png}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20\%,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb-style pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (denoting that only the offset position is missing).
        The dark vertical strip near $l \sim$125\deg
        is an uninteresting consequence of the tiling pattern,
        which overlaps in the middle of the survey area by design.}
        \label{fig:footprint}
\end{figure*}

Fig.~\ref{fig:footprint} shows the footprin
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8\% of the survey area --
are predominantly located towards the Galactic anti-centre 
at $l > 120^o$.
Fields at these longitudes can only be accessed from La Palma 
in the months of November-December,
which is when the weather and seeing conditions are often poor
at the INT and observing attempts have failed repeatedly.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is available.

@@

All raw data obtained with the INT were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archival.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ survey programme carried out 
between 1998 and 2003 after the WFC was commissioned.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A detailed description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and much of the source code is available 
on line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

The reduced images are then stored in multi-extension FITS files
with a primary header describing the characteristics
(position, filter, exposure time, etc.) 
and four image extensions 
corresponding to each of the four CCDs.
Source detection and characterisation is then carried out 
using the \textsc{imcore} tool \citep{Irwin1985,Irwin1997}.
The flux of each source is measured using both
the peak pixel height (i.e. a square 0.33$\arcsec\times$0.33\arcsec\ aperture)
as well as a series of circular apertures of increasing diameter 
(1.2\arcsec, 2.3\arcsec, 3.3\arcsec, 4.6\arcsec\ and 6.6\arcsec).

The local background levels are estimated 
by computing the sigma-clipped median
flux in a grid of 64$\times$64 pixels (21\arcsec$\times$21\arcsec)
across the image,
which is then interpolated to obtain an estimate 
of the background level at each pixel.
These sky levels are subtracted from the aperture photometry and
-- when required --
a deblending routine is applied which attempts to remove
the contamination from any nearby sources.
Whilst this approach works very well 
across the vast majority of the survey area,
the Galactic Plane unavoidably contains crowded regions 
with large numbers of overlapping sources
or rapidly spatially-varying nebulosity,
in which case aperture photometry must always be interpreted 
with caution.
In \S\ref{sec:catalogue} we will explain that objects
which suffer from this problem 
are flagged in the catalogue using the \emph{deblend} warning flag.

Finally, an astrometric solution is determined
based on the Two-Micron All Sky Survey (2MASS) point source catalogue \citep{Skrutskie2006},
which itself is calibrated 
in the International Celestial Reference System (ICRS).
A provisional photometric calibration is also provided 
based on the average zeropoint
determined from a set of standard stars observed in the same night.
Sources are classified morphologically
-- stellar, extended or noise --
based on the curve-of-growth determined
from measuring the source intensity in a series of growing apertures.
Finally, the resulting source detection tables are also stored 
in multi-extension FITS files.

At the time of preparing DR2,
the CASU pipeline had processed
74\,195 IPHAS exposures 
in which a total of 1.9~billion \emph{candidate sources} were detected 
at the sensitive default detection level of 1.25\,$\sigma$.
This unavoidably includes spurious detections, artefacts and
duplicate detections,
in \S\ref{sec:catalogue} we will explain
how these have been removed or flagged in the final catalogue.
The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a reliable source catalogue which is 
(i) quality-controlled,
(ii) homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which are explained next.

@@

Observing time for IPHAS was obtained
on a semester-by-semester basis
through the traditional time allocation committees 
of the Isaac Newton Group of telescopes,
which are invariably over-subscribed.
For this reason, we attempted to utilise 
\emph{all} the nights allocated to IPHAS,
even those which were partially or entirely non-photometric
or otherwise affected by technical problems 
(e.g. electronic noise or telescope tracking problems).
Any unsuitable data that were taken as a result
of this strategy has subsequently been flagged and removed
using a series of seven quality criteria,
which ensure a reliable and homogeneous level of quality
across the data release:

\begin{figure}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/depth_r.pdf} 
    \end{minipage}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/depth_i.pdf} 
    \end{minipage}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/depth_h.pdf} 
    \end{minipage}
    \caption{Distribution of the 5$\sigma$ limiting magnitude
             across all quality-approved survey fields
             for $r$ (top), $i$ (middle) and \ha\ (bottom).
             Fields with a limiting magnitude brighter than
             20th ($r$) or 19th (\ha, $i$) were rejected
             from the data release.
             The $r$-band depth is most sensitive 
             to the presence of the moon above the horizon, 
             which is evidenced by the wide and bi-model shape
             of its distribution.}
    \label{fig:depth}
\end{figure}

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was worse than 20th magnitude in the $r$-band
or worse than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations fared significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields,
showing a mean depth of 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended 
most strongly on the presence of the moon,
which was above the horizon during 62\% 
of the observations and explains the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight,
and their magnitude limit distributions
are hence more narrow
(middle and bottom panels in Fig.~\ref{fig:depth}).

The minimum depth criterion
has led us to exclude 9\% of the pipelined data,
which were often obtained under cloudy conditions.
We note that some of the excluded data may nevertheless be useful
for e.g. time-domain studies of bright stars.
The detection tables of the discarded data are made
available through our website for this reason (www.iphas.org),
but are ignored in what follows.

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation of the point spread function.
It is expected to be zero (circular) across the field 
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.
The mean ellipticity across a field in the IPHAS data set 
is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
For this reason, we rejected exposures in which the mean ellipticity
across the detector exceeded $e > 0.3$,
which is when the photometric measurements delivered by the pipeline
were found to become degraded.
Only 0.4\% of the exposures were discarded on this basis.

(3) \emph{Seeing.} 
The survey originally aimed to obtain data 
at seeing better than 1.7~arcsec.
This target is currently attained across 86\% of the footprint,
in particular at early longitudes,
e.g. 92\% of the fields at $l<120^\circ$ are better than 1.7~arcsec.
Fig.~\ref{fig:seeing} presents the distribution
of the mean seeing for all the quality-approved fields.
We find a median value of 1.1~arcsec in $r$, \ha\
and 1.0~arcsec in $i$.
In the $r$-band, 90\% of the data is better than 1.5~arcsec,
and 10\% is better than 0.8~arcsec.
To improve the area covered by this data release,
we have decided to include the small fraction of data
that was obtained under seeing up to 2.5 arcsec,
so that only 1\% of the pipelined exposures
had to be excluded due to poor seeing.
In \S\ref{sec:catalogue} we will explain
that the photometry listed in the catalogue
is predominantly based on the images with the
best-available seeing, and that the seeing estimate
is included as a column in the catalogue.

>>>>

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering the Galactic latitude range $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
uniformly-calibrated source catalogue
derived from the survey,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the 1800 square-degree footprint.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
and to a mean $5\sigma$-depth of 
21.2 ($r$), 20.0 ($i$) and 20.3 ($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the new 98-column source catalogue.
We show that the global calibration is accurate to
0.03~mag (rms) 
%by validating against the Sloan Digital Sky Survey (SDSS),
and provide a series of recommended quality criteria
which can be used to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select candidate \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure of the disk of the Milky Way.
This large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights 
at the Isaac Newton Telescope (INT) in La Palma --
aims to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century \citep[see][]{Kohoutek1999}. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
IPHAS can expand
the limited bright samples of Galactic emission line objects 
previously available into larger, deeper, and more statistically-robust samples that will 
better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the IPHAS Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
%new asymptotic giant branch stars \citep{Wright2008};
new classical Be stars \citep{Raddi2013};
along with discoveries of new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Corradi2011, Viironen2011}
and new supernova remnants \citep{Sabin2013}.

Over the years it has become apparent that the legacy of IPHAS 
will reach beyond these traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$, $i$ broadband photometry 
alongside H$\alpha$ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey’s unique $(r-\ha,\ r-i)$ colour-colour
diagram has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband H$\alpha$ 
alongside $r, i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys.
For such science applications to succeed, however, 
it is vital that the imaging data are transformed 
into a uniformly-calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

When the initial data release was created in late 2007,
just over $\sim$half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release 
that takes the coverage up to 92 percent of the survey area 
and includes a uniform calibration.
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new global photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

In \S\ref{sec:observations} we start by recapitulating the key points
of the survey observing strategy.
In \S\ref{sec:reduction} we describe the data reduction
and quality control procedures.
In \S\ref{sec:calibration} we explain the uniform re-calibration, in which
we draw upon the AAVSO Photometric All-Sky Survey (APASS)
and test our results against the Sloan Digital Sky Survey (SDSS).
In \S\ref{sec:catalogue} we explain how the source catalogue was compiled.
In \S\ref{sec:discussion} we discuss the properties of the catalogue
and in \S\ref{sec:demonstration} we demonstrate
the scientific exploitation of the colour/magnitude diagrams.
Finally, in \S\ref{sec:dataaccess} we discuss access
to the catalogue, an online library of reduced images and relevant source code,
The paper ends with conclusions in \S\ref{sec:conclusions} where we also outline
our future ambitions.

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release, we provide a reminder of some key points.
IPHAS is an imaging survey of the Galactic Plane north of the celestial equator, 
from which photometry in Sloan $r, i$ 
is extracted along with narrowband H$\alpha$. 
It is carried out using the Wide Field Camera (WFC) 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive optical CCD photometry
of point sources in the Galactic Plane, 
and does so down to a limiting magnitude of $\sim$20.
The IPHAS footprint spans a box 
of roughly 180 by 10 degrees, 
spanning Galactic latitudes $-5^{\circ} < b < +5^{\circ}$ 
and longitudes $30^{\circ} < l < 215^{\circ}$.

The WFC is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29~deg$^2$.
To cover the entire Northern Plane with some overlap,
the survey area was divided into 7635 telescope pointings.
Each of these pointings is accompanied by an offset position
displaced by $+$5 arcmin in declination 
and $+$5 arcmin in right ascension,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks. 
Hence, the basic unit of observation hence
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within, typically, an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
Altogether, the survey contains 15\,270 fields grouped into 7\,635 field pairs.
To achieve the desired survey depth of 20th magnitude or fainter, 
the filter exposure times were set at 120 sec (narrowband \ha), 
30 sec ($r$) and 10 sec ($i$)
in the majority of the survey observations.\footnote{In 2003 
the $r$-band exposure time was 10~sec instead of 30~sec,
and since Oct 2010 the $i$-band exposure time 
has been increased from 10~sec to 20~sec.}

Data-taking began in the second half of 2003, 
and every field had been observed at least once by the end of 2008.
At that time only 76\% of the field pairs 
satisfied our minimum quality criteria, however.  The problems affecting
the 24\% falling below survey standard were, most commonly: variable cloud cover; 
poor seeing; technical faults (the quality criteria will be detailed in the 
next section). 
Since then, a programme of repeat observations has been in place 
to improve data quality. 
As a result, 92\% of the survey footprint
now benefits from quality-approved data.
The most recent observations included in this release
were obtained in November 2012.

\begin{figure*}
        \includegraphics[width=1\linewidth]{./plots/footprint/footprint_small.png}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20\%,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (denoting that only one offset is missing).
        The dark vertical strip near $l \sim$125\deg
        is an arbitrary consequence of the tiling pattern,
        which was populated starting from 0h in right ascension.}
        \label{fig:footprint}
\end{figure*}

Fig.~\ref{fig:footprint} shows the footprin
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8\% of the survey area --
are predominantly located towards the Galactic anti-centre 
at $l > 120^o$.
Fields at these longitudes are mainly accessed from La Palma 
in the months of November-December,
which is when the La Palma weather and seeing conditions are often poor
forcing many (unsuccessful) repeat observations.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is available.

@@

All raw IPHAS data were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archival.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ survey programme carried out 
between 1998 and 2003 after the WFC was commissioned.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A detailed description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and much of the source code is available 
on line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

The reduced images are then stored in multi-extension FITS files
with a primary header describing the characteristics
(position, filter, exposure time, etc.) 
and four image extensions 
corresponding to each of the four CCDs.
Source detection and characterisation is then carried out 
using the \textsc{imcore} tool \citep{Irwin1985,Irwin1997}.
The flux of each source is measured using both
the peak pixel height (i.e. a square 0.33$\arcsec\times$0.33\arcsec\ aperture)
and a series of circular apertures of increasing diameter 
(1.2\arcsec, 2.3\arcsec, 3.3\arcsec, 4.6\arcsec\ and 6.6\arcsec).

The local background levels are estimated 
by computing the sigma-clipped median
flux in a grid of 64$\times$64 pixels (21\arcsec$\times$21\arcsec)
across the image,
which is then interpolated to obtain an estimate 
of the background level at each pixel.
These sky levels are subtracted from the aperture photometry and
-- when required --
a deblending routine is applied to remove
the contamination from any nearby sources.
This approach works very well 
across the vast majority of the survey area.  Nevertheless,
the Galactic Plane contains crowded regions 
with large numbers of overlapping sources
or rapidly spatially-varying nebulosity,
where aperture photometry will be compromised by frequent blending
or poor background subtraction.
In \S\ref{sec:catalogue} we will explain how affected sources are 
flagged. 
%we will explain that objects affected in this way 
%are flagged in the catalogue using the \emph{deblend} warning flag.

Finally, an astrometric solution is determined
based on the Two-Micron All Sky Survey (2MASS) point source catalogue \citep{Skrutskie2006},
which itself is calibrated 
in the International Celestial Reference System (ICRS).
A provisional photometric calibration is also provided 
based on the average zeropoint
determined from sets of standard stars observed within the same night.
Sources are classified morphologically
-- stellar, extended or noise --
based on the curve-of-growth determined
from measuring the source intensity in a series of growing apertures.
Finally, the resulting source detection tables are also stored 
in multi-extension FITS files.

At the time of preparing DR2,
the CASU pipeline had processed
74\,195 IPHAS exposures 
in which a total of 1.9~billion \emph{candidate sources} were detected 
at the sensitive default detection level of 1.25\,$\sigma$.
This total inevitably includes spurious detections, artefacts and
duplicate detections,
in \S\ref{sec:catalogue} we will explain
how these have been removed or flagged in the final catalogue.
The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a reliable source catalogue which is 
(i) quality-controlled,
(ii) homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which are explained next.

@@

Observing time for IPHAS was obtained
on a semester-by-semester basis
through the open time allocation committees 
of the Isaac Newton Group of telescopes,
which are invariably over-subscribed.
For this reason, we attempted to utilise 
\emph{all} the nights allocated to IPHAS,
even those which were partially or entirely non-photometric
or otherwise affected by technical problems 
(e.g. electronic noise or telescope tracking problems).
Any unsuitable data that were taken as a result
of this strategy have subsequently been flagged and removed
using seven quality criteria,
which ensure a reliable and homogeneous level of quality
across the data release:

\begin{figure}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/depth_r.pdf} 
    \end{minipage}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/depth_i.pdf} 
    \end{minipage}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/depth_h.pdf} 
    \end{minipage}
    \caption{Distribution of the 5$\sigma$ limiting magnitude
             across all quality-approved survey fields
             for $r$ (top), $i$ (middle) and \ha\ (bottom).
             Fields with a limiting magnitude brighter than
             20th ($r$) or 19th (\ha, $i$) were rejected
             from the data release.
             The $r$-band depth is most sensitive 
             to the presence of the moon above the horizon: 
             this is the reason for the wide, bi-modal character
             of its distribution.}
    \label{fig:depth}
\end{figure}

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was worse than 20th magnitude in the $r$-band
or worse than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields,
showing a mean depth of 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended 
most strongly on the presence of the moon,
which was above the horizon during 62\% 
of the observations.  The great range in sky brightness this
produced is the reason for the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).

The minimum depth criterion
led us to exclude 9\% of the pipelined data,
which had typically been obtained under cloudy conditions.
We note that some of the excluded data may nevertheless be useful
for e.g. time-domain studies of bright stars.
The detection tables of the discarded data are made
available through our website for this reason (www.iphas.org),
but are ignored in what follows.

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation of the point spread function.
It is expected to be zero (circular) across the field 
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.  Indeed, it is worth noting
that IPHAS data have been collected from unguided exposures that rely
entirely on the INT's tracking capability.
The mean ellipticity measured 
%across a field 
from exposures forming this release
%in the IPHAS data set 
is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
To exclude these, we rejected exposures in which the mean ellipticity
across the detector exceeded $e > 0.3$. We have also found that, as
this threshold is breached, the photometric measurements delivered by 
the pipeline become degraded.
Only 0.4\% of the exposures were discarded on this basis.

(3) \emph{Seeing.} 
The original survey goal was to obtain data 
at seeing better than 1.7~arcsec.
This target is currently attained across 86\% of the footprint,
in particular at lower Galactic longitudes,
e.g. 92\% of the fields at $l<120^\circ$ are better than 1.7~arcsec.
Fig.~\ref{fig:seeing} presents the distribution
of the mean seeing for all the quality-approved fields.
We find a median value of 1.1~arcsec in $r$, \ha\
and 1.0~arcsec in $i$.
In the $r$-band, 90\% of the data is better than 1.5~arcsec,
and 10\% is better than 0.8~arcsec.
To increase the sky area covered by this data release,
we have decided to include the small fraction of data
obtained under seeing of up to 2.5 arcsec.  This limited
exclusion on grounds of poor seeing to only 1\%\ of the pipelined 
exposures. 
In \S\ref{sec:catalogue} we will explain
that the photometry listed in the catalogue
is predominantly based on the images with the
best-available seeing, and that the seeing estimate
is included as a column in the catalogue.
